---
title: "Lists - In Class Exercises"
author: "ERICA GALVAN"
date: "DATA 2401"
output: html_document
---

# Exercises: 

Make a list that contains the information on a single playing card. Suit, face, and value. 

```{r}
playing_card <- list(
  suit = c("Hearts"),
  face = c("Queen"),
  value = c("12")
)
```


Make a second list where each element contains all the options for suit, face, and value, respectively. 

```{r}
#creating new list 
cards <- list(
  suit = "spades", "hearts", 
  face = "ace", "queen", "jack", "king", 
  value = 0, 11, 12, 13
)
```


If `Newlist <- list(a=1:10, b="Good morning", c="Hi")`, write an R statement that will add 1 to each element of the first vector in Newlist.

```{r}
Newlist <- list(a=1:10, b="Good morning", c="Hi")

Newlist [["a"]]+1
#or
Newlist$a +1
```

If `b <- list(a=1:10, c="Hello", d="AA")`, write an R expression that will give all elements of the first vector of b, except the second, and write another that returns only elements of `a` that are greater than 7.

```{r}
b <- list(a=1:10, c="Hello", d="AA")
#these are both the same *just practicing 
b [["a"]]
b$a
#checking which elements that are greater than 7
b$a > 7

#return the elements if they were true 
b$a [b$a >7]
```

Let `x <- list(a=5:10, c="Hello", d="AA")`, write an R statement to add a new item z = "NewItem" to the list x.

```{r}
x <- list(a=5:10, c="Hello", d="AA")

x$z <- "New Item"
```
